Cheminformatics: The Computer Science of Chemical Discovery

### 2011-11-18 paper.tex

\additionalauthors{Additional authors:
Andreas Bender (University of Cambridge, email: {\texttt{andreas.bender@cantab.net}}),
Jean-Loup Faulon (University of Evry, email: {\texttt{Jean-Loup.Faulon@issb.genopole.fr}}),
Janna Hastings (European Bioinformatics Institute, Cambridge, UK, email: {\texttt{janna.hastings@gmail.com}}),
Noel O'Boyle (University College Cork, Cork, Ireland, email: {\texttt{baoilleach@gmail.com}}),
John Overington (European Bioinformatics Institute, Cambridge, UK, email: {\texttt{jpo@ebi.ac.uk}}),
Herman Van Vlijmen (Tibotec, Beerse, Belgium, email: {\texttt{hvvlijme@its.jnj.com}}), and
Egon Willighagen (Karolinska Institutet, Stockholm, Sweden, email: {\texttt{egon.willighagen@ki.se}})
.}
\date{25 June 2011}

\maketitle
\begin{abstract}
  One of the most prominent success stories in all the sciences over
  the last decade has been the advance of bioinformatics: The
  interdisciplinary collaboration between computer scientists and
  molecular biologists led as one of its major accomplishments to the
  sequencing of the human genome, where researchers connected
  biological concepts to the theory of string algorithms. However,
  despite this great success, few computer scientists are familiar
  with a related discipline: cheminformatics, the use of computers to
  represent the structures of small molecules and analyze their
  properties.  Cheminformatics has wide applicability, from the
  discovery of novel drug candidates to the optimization of certain
  physicochemical properties of small molecules. Until recently, much
  of the the data and some of the techniques employed in the
  cheminformatics field have been closely guarded secrets of companies
  whose financial success depended on being the first to produce the
  new therapeutic molecules.  Only within the last decade -- and as an
  effect of a change in mindsets, government-mandated policy changes,
  and, not least, because of chemists volunteering their time
  for an Open Science ``movement'' -- have researchers gained access to
  freely available software packages and databases of tens of millions
  of chemicals. Academic chemists now confront a variety of unsolved
  algorithmic problems that could not have been tackled a decade ago,
  but whose solutions are critical to research ranging from
  determining the behavior of small molecules in biological pathways,
  to finding therapies for rare and neglected diseases.
\end{abstract}

@@

Novel technologies in the life sciences churn out information at an
ever increasing rate -- public data stores such as the one at the
European Bioinformatics Institute (EBI) contain on the order of 10
petabytes of biological information.  While biological information has
been publicly available for decades, the same has not been true for
chemical information, until recently. It was not until 2004 that a
large public small molecule structure repository (PubChem) became
freely available. This was followed by other databases. While much of
the foundational algorithms of cheminformatics have been described
since the 1950s, publicly available software implementing many of
these algorithms have only been accessible in the last ten to fifteen
years \cite{faulon2010}.

@@

Cheminformatics comprises different areas, which broadly speaking can
be divided into three fields: \emph{capturing data} (using lab
notebooks or potentially using formats such as the Chemical Markup
Language for publications); \emph{storing data} (designing database
schemas, devising ontologies) and \emph{mining data} (such as for the
prediction of biological activities of compounds).  However, the type
of information being analyzed is different from biological
information: while bioinformatics often deals with sequences, the
domain of cheminformatics is chemical structures. In the case of the
former, information can very frequently be represented as
one-dimensional strings, which are relatively easy to handle
computationally. In the case of the latter however, chemical
structures may possess rings, branches as well as multiple valid
representations of the same structure (such as \emph{tautomers}, where
hydrogen atoms can be positioned in different places of the
molecules), potentially giving rise to ambiguities. Hence, chemical
structures are often reported to be more difficult to standardize and
algorithms dealing with them must take this into account.

@@

\begin{figure}
\centering
\includegraphics[height=2in]{chemical-structures.png}
\caption{A hierarchy of chemical structure representations, increasing
  in information content from left to right and top to bottom. A
  chemists traditional 2D depiction (A), a labeled graph (B), a 3D
  conformation (C) and a surface represention (D). All show differing
  (and valid) aspects of the same molecule and are suited for
  differing purposes. The choice of representation is guided by the
  intended use of the structure data.}
\label{figure:chemical-structures}
\end{figure}

>>>>

\additionalauthors{Additional authors:
Andreas Bender (University of Cambridge, email: {\texttt{andreas.bender@cantab.net}}),
Jean-Loup Faulon (University of Evry, email: {\texttt{Jean-Loup.Faulon@issb.genopole.fr}}),
Janna Hastings (European Bioinformatics Institute, Cambridge, UK, email: {\texttt{hastings@ebi.ac.uk}}),
Noel O'Boyle (University College Cork, Cork, Ireland, email: {\texttt{baoilleach@gmail.com}}),
John Overington (European Bioinformatics Institute, Cambridge, UK, email: {\texttt{jpo@ebi.ac.uk}}),
Herman Van Vlijmen (Tibotec, Beerse, Belgium, email: {\texttt{hvvlijme@its.jnj.com}}), and
Egon Willighagen (Karolinska Institutet, Stockholm, Sweden, email: {\texttt{egon.willighagen@ki.se}})
.}
\date{25 June 2011}

\maketitle
\begin{abstract}
  One of the most prominent success stories in all the sciences over
  the last decade has been the advance of bioinformatics: the
  interdisciplinary collaboration between computer scientists and
  molecular biologists led as one of its major accomplishments to the
  sequencing of the human genome, where researchers connected
  biological concepts to the theory of string algorithms. However,
  despite this great success, few computer scientists are familiar
  with a related discipline: cheminformatics, the use of computers to
  represent the structures of small molecules and analyze their
  properties.  Cheminformatics has wide applicability, from the
  discovery of novel drug candidates to the optimization of certain
  physicochemical properties of small molecules. Until recently, much
  of the data and some of the techniques employed in the
  cheminformatics field have been closely guarded secrets of companies
  whose financial success depended on being the first to produce
  new therapeutic molecules.  Only within the last decade -- as an
  effect of a change in mindsets, government-mandated policy changes,
  and, not least, as a result of chemists volunteering their time
  for an Open Science ``movement'' -- have researchers gained access to
  freely available software packages and databases of tens of millions
  of chemicals. Academic chemists now confront a variety of unsolved
  algorithmic problems that could not have been tackled a decade ago,
  but whose solutions are critical to research ranging from
  determining the behavior of small molecules in biological pathways,
  to finding therapies for rare and neglected diseases.
\end{abstract}

@@

Novel technologies in the life sciences churn out information at an
ever increasing rate -- public data stores such as that of the
European Bioinformatics Institute (EBI) contain in the order of 10
petabytes of biological information.  While biological information has
been publicly available for decades, the same has not been true for
chemical information, until recently. It was not until 2004 that a
large public small molecule structure repository (PubChem) became
freely available. This was followed by other databases. While much of
the foundational algorithms of cheminformatics have been described
since the 1950s, publicly available (open source) software implementing many of
these algorithms has only been accessible in the last ten to fifteen
years \cite{faulon2010}.

@@

Cheminformatics comprises different areas, which broadly speaking can
be divided into three fields: \emph{capturing data} (using lab
notebooks or potentially using formats such as the Chemical Markup
Language for publications); \emph{storing data} (designing database
schemas, devising ontologies) and \emph{mining data} (such as for the
prediction of biological activities of compounds).  However, the type
of information being analyzed is different from biological
information: while bioinformatics often deals with sequences, the
domain of cheminformatics is chemical structures. In the case of the
former, information can very frequently be represented as
one-dimensional strings, which are relatively easy to handle
computationally. In the case of the latter however, chemical
structures may possess rings, branches as well as multiple valid
representations of the same molecule (such as \emph{tautomers}, where
hydrogen atoms can be positioned in different places of the
molecules), potentially giving rise to ambiguities. Hence, chemical
structures are often reported to be more difficult to standardize and
algorithms dealing with them must take this into account.

@@

\begin{figure}
\centering
\includegraphics[height=2in]{chemical-structures.png}
\caption{A hierarchy of chemical structure representations, increasing
  in information content from left to right and top to bottom. A
  chemist's traditional 2D depiction (A), a labeled graph (B), a 3D
  conformation (C) and a surface represention (D). All show differing
  (and valid) aspects of the same molecule and are suited for
  different purposes. The choice of representation is guided by the
  intended use of the structure data.}
\label{figure:chemical-structures}
\end{figure}
